\documentclass{revtex4}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{braket}
\usepackage{mathtools}
\usepackage{textcomp}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{color}

\begin{document}
\title{Hartree-Fock Stability Theory and its Relation to Correlation Theories}
\author{Evan Curtin}
\maketitle


\section{Background}

    The goal of electronic structure theory is to solve the electronic Schr{\"o}dinger equation
    for a molecule or solid state system. The prevailing approach in \emph{ab initio} quantum 
    chemistry is to first compute a self-consistent single-determinant Hartree-Fock (HF) solution 
    which will serve as a reference for many-body perturbative (MP) or coupled-cluster (CC) 
    approaches.  
    Implicit in these methods is the assumption that the reference wave function will be 
    sufficiently close to the exact solution of the Schr{\"o}dinger equation to converge to the 
    correct 
    answer in a computationally feasible manner. Both the MP and CC approaches have the attractive 
    quality that they are systematically improvable, i.e. more computational power can be used on a 
    problem to increase the accuracy of the resulting solution. These so-called post-Hartree-Fock 
    methods are able to recover more of the total electronic energy by accounting for the 
    \emph{correlation energy} in the system. Traditionally, the correlation energy is defined by 
    the difference between the exact energy and the Hartree-Fock energy \cite{Shavitt2009}
    \begin{equation}\label{correlation_energy}
      E_{corr} = E_{exact} -  E_{HF}
    \end{equation}
    In many cases the correlation energy is relatively small and the Hartree-Fock reference 
    captures a significant portion of the exact total energy. In such cases, the prevailing 
    approach is usually successful and has only implementation issues for larger systems. On the 
    other hand there are systems which are not well treated by this approach, which are often 
    called "strongly correlated". Examples of such systems are those with degenerate or 
    near-degenerate HOMO-LUMO gaps, metallic solids or molecules close to breaking bonds. 
    Oftentimes for such systems the traditional Hartree-Fock solutions are not even qualitatively 
    correct, implying inadequacies in either the single-reference or mean-field approach. 
      
      It is often neglected, however, that the usual way for computing a Hartree-Fock solution in 
      practice
    is to use a restricted form of the single determinantal wavefunction. In the Restricted 
    Hartree-Fock (RHF), the wave function is restricted to be an eigenfunction of the $\hat{S^2}$ 
    and $\hat{S_z}$ 
    operators, while in the Unrestricted variant (UHF), the wave function need only be an 
    eigenfunction 
    of the $\hat{S_z}$. There is yet another lesser-known variation of the theory, known as General 
    Hartree-Fock (GHF), which assumes no spin-symmetries in the wave function. This allows for 
    greater
    variational flexibility at the cost of "broken symmetry". Together these three levels of 
    Hartree-Fock theory form a heirarchy of increasing parameter space within which to minimize the 
    total energy. Thus the term "Hartree-Fock Energy" is inherently ambigious; this fact calling 
    into question commonly held ideas about correlation and Hartree-Fock theory. Take for example 
    the dissociation curve of H$_2$ in the RHF and UHF approximations, compared with the exact 
    solution obtained by full configuration-interaction in the complete basis set limit. 
    
    \textcolor{red}{H2 curve pic}
    
    The correlation energy for the RHF reference case is dramatically larger than that for the UHF 
    reference. In other words, \emph{the correlation energy is, by definition, inherently coupled 
    to the 
    Hartree-Fock reference}. The given example is well understood, but it brings to mind a more 
    general idea that there may be more cases where apparently large correlation energies are due 
    to simply 
    using and inadequate level of Hartree-Fock theory.  
    
    
    
 
    The Hartree-Fock procedure has been implemented the quantum many-body problem since its inception and remains the basis for many more advanced techniques even today. In general, the eigenstates of the Hartree-Fock Hamiltonian are solved self-consistently. However, this procedure ensures only that the solution is stationary with respect to the determined orbitals. A method for determining the stability of a Hartree-Fock solution was proposed by Thouless in 1960\cite{Thouless1960}. The condition for stability of a Hartree-Fock solution is equivalent to the conditions for unstable (complex frequency) many-body oscillations within the Random Phase Approximation (RPA). The condition was rederived into the expression familiar to quantum chemists by Čížek and Paldus in 1967\cite{Cizek1967}. Furthermore, the stability equations factorize depending on the symmetry of the Hartree-Fock eigenfunctions. To this end, Seeger and Pople outlined a hierarchical approach to systematically evaluate the stability of HF states in the restricted, unrestricted and generalized Hartree-Fock procedures\cite{Seeger1977}. Recently, the method has been used to aid the in search for the lowest energy Unrestricted Hartree-Fock (UHF) solutions in molecules, as well as the General Hartree-Fock (GHF) solutions in geometrically frustrated hydrogen rings which cannot conform even to the UHF scheme \cite{Pulay2016}\cite{Goings2015}.
    
    The presence of GHF solutions to the Homogeneous Electron Gas which have lower energy than the RHF solutions was proven in the landmark paper by Overhauser\cite{Overhauser1962}. Later still, the ground state energies of the electron gas were found to great accuracy by Ceperley and Alder \cite{Ceperley1980}. In the past few years, phase diagrams have been determined for the HEG in 2 and 3 dimensions\cite{Delyon2008}\cite{Bernu2011}\cite{Baguet2013}. In all cases the phase diagrams are made by computing the energies of the polarized and unpolarized states, and comparing them. The present work focuses on directly calculating the stability of the paramagnetic homogeneous electron gas as a function of electron density. It should be emphasized that instability of the Hartree-Fock solution is a sufficient, but not necessary, condition for the presence of a symmetry-broken Hartree-Fock solution of lower energy to exist. The technique may be used to guide a search for the lowest energy Hartree-Fock solution by following the instability eigenvectors downhill in energy. Such an approach will be attempted in a future paper. 
    
\section{Methods}    
	
	Starting with the Jellium Hamiltonian, 
	\begin{equation}\label{hamiltonian}
		\mathbf{H} = ....
	\end{equation}
	
	Solutions to the Hartree-Fock equations in $D$ dimensions are plane waves of the form, 
	\begin{equation}\label{planewave}
		\phi_{\vec{k} \sigma} =
		   \frac{1} { \sqrt{L ^ D} } e ^ {i \vec{k} \dot \vec{x}}.
	\end{equation}
	Choosing a plane wave basis with periodic boundary conditions is therefore natural, and results in the two electron repulsion integrals being analytic with the following forms, (eq. 12 of \cite{Delyon2008} and p. 16 of \cite{Guiliani2005})
    
    \begin{subequations}
    	\begin{align}
    	\braket{\vec{k}, \vec{k}' | \vec{k}'', \vec{k}'''} 
    	  \stackrel{ \text{2D, 3D} }{=}&
    	\begin{cases} 
    	\frac{\pi} {\Omega} \frac{ 2^{D-1} } { | \vec{k} - \vec{k}'' | ^ {D-1} } 
    	& \vec{k}''' = \vec{k} + \vec{k}' - \vec{k}'' + \vec{G} \textbf{ and } | \vec{k} - \vec{k}''| \neq 0 \\
    	0 
    	& \text{else}
    	\end{cases}
    	\\ \nonumber \\
    	\braket{k, k' | k'', k'''} \stackrel{\text{1D}}{=}&
    	\begin{cases} 
    	e ^ { |k-k''|^2a^2 } \text{Ei} (-|k - k''|^2 a^2)\text{ ;}
    	& k''' = k + k' - k'' + \vec{G} \textbf{ and } |k - k''| \neq 0 \\
    	0\text{ ;} 
    	& \text{else}
    	\end{cases}
    	\end{align}
    \end{subequations}
    where $\Omega$ is the direct lattice volume and $\vec{G}$ is a reciprocal lattice vector of the system. Ei(x) denotes the exponential integral function and the parameter $a$ in 1D is the radius of a cylinder that approximates a 1D system to circumvent divergence of the integral in 1 dimension. The orbital energies are given 
    by 
    
    \begin{equation}\label{eq:hf_orb_energy}
    \epsilon_{\vec{k},\sigma}=
    \frac{\hbar^2\vec{k}^2}{2m} - \sum\limits_{\vec{k}}^{|\vec{k}|< k_f}n_{\vec{k}\sigma}\braket{\vec{k}, \vec{k}' |\vec{k}', \vec{k}}
    \end{equation}
    
    Where $n_{\vec{k}\sigma}$ is the occupation number of the state with momentum $\vec{k}$ and spin $\sigma$ (p. 15 of \cite{Guiliani2005}). 

    \subsection{Hartree-Fock Stability}
    The HF stability condition was presented by Thouless \cite{Thouless1972} ($1^{st}$ Ed. 1960) and is the presence of imaginary electronic oscillation frequencies in the random phase approximation (RPA). Equivalently, instability of a Hartree-Fock solution is borne out as a negative eigenvalue of the electronic Hessian (static density-matrix--density-matrix response function). For a paramagnetic HF solution, the equations factorize into the matrices known to chemists as the singlet and triplet instability matrices \cite{Dunning1967}\cite{Seeger1977} \textcolor{red}{show this here? appendix?}
    ,
    \begin{equation}\label{eq:1,3H}
    {}^{1,3}\bf{H'}=
    \begin{bmatrix}
    {}^{1,3}\bf{A'} & {}^{1,3}\bf{B'} \\
    \left({}^{1,3}\bf{B'}\right)^* & \left({}^{1,3}\bf{A'}\right)^* \\
    \end{bmatrix},
    \end{equation}
    where the $1,3$ denote singlet and triplet instabilities, respectively. $\bf{A}$, $\bf{B}$ have dimension $N_{occ}\times N_{vir}$ and are defined as follows:
    \begin{subequations}
    	\begin{align}
    	{}^{1}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} + 2\braket{aj|ib}-\braket{aj|bi}\\
    	{}^{3}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} - \braket{aj|bi}\\
    	{}^{1}{B'}_{i\rightarrow a, j\rightarrow b} &= 2\braket{ab|ij}-\braket{ab|ji}\\
    	{}^{3}{B'}_{i\rightarrow a, j\rightarrow b} &= -\braket{ab|ji}
    	\end{align}
    \end{subequations}
    The matrix given in Equation \ref{eq:1,3H} represents the case of complex RHF stability in the 
    space of complex RHF space (${}^{1}\mathbf{H}$ - internal instability) and in the space of 
    complex UHF space (${}^{3}\mathbf{H}$ - triplet instability). The lowest eigenvalue of both of 
    these matrices will reveal the stability of the RHF solution. If they are both positive, the 
    solution is stable with respect to symmetry breaking, while negative eigenvalues in either 
    indicate instability with respect to either triplet or singlet perturbations (or both). We 
    mention here that an eigenvalue of 0 does not indicate instability, and this case is discussed 
    in detail by Cui et al \cite{Cui2013}.
    
    \section{Results}
    
    

\section{References}
\bibliography{../prelim_references}

\end{document}